---
title: "insects"
author: "Pete Nelson"
format: html
editor: visual
---

# intro

Explore Rosie's data to see what monitoring programs and sampling techniques seem adequate for studying the occurrence of insects. Then, to see how effectively these have been applied to the project areas that we're interested in. I'll start by loading some libraries and then looking at some of the data files that Rosie's provided (no

```{r}
#| label: data
#| echo: false

library(here)
library(janitor)
library(tidyverse)
library(brms)
library(patchwork)

load(here("data/AllWetlandBugs.RData")) # AllBugs
load(here("data/AllWetlandBugs_2010onwards.RData")) # Allbugs_Oct2025
load(here("data/Bugs_allfilters.RData"))
load(here("data/Insects.RData"))
glimpse(AllBugs) # 427,198 x 38
glimpse(Allbugs_Mar2026) # 2,266,598 x 40
glimpse(Bugs_allfilters) # 874,790 x 47
glimpse(Insects) # 7,601 x 27

```

I'm unfamiliar with all of the sites, most of the monitoring programs, and many of the sampling techniques (or at least what their codes refer to). The `Insects` data file is limited to chironomids and all other insects, so I'm going to poke around in `Bugs_allfilters` data (though it may well be that that's the best taxonomic resolution we can hope for). This may be the wrong place to start; if so, I'm hoping it'll be relatively easy to correct later.

# data exploration

Add `brms` and `patchwork` to the library block if not already loaded.

I want to see how well different sampling programs ('Source') and sampling techniques ('TowType') work for insects.

```{r}
#| label: sampling-efficacy

Bugs_allfilters %>%
  filter(Class == "Insecta") %>%
  group_by(Source, TowType, SampleID) %>%
  summarise(sample_CPUE = sum(CPUE, na.rm = TRUE), .groups = "drop") %>% # one row per tow (sample unit)
  group_by(Source, TowType) %>%
  summarise(
    n_samples = n_distinct(SampleID),
    # tows that caught zero insects
    n_zero_samples = sum(sample_CPUE == 0), 
    # handy for the expected hurdle question
    prop_zero = mean(sample_CPUE == 0), 
    # mean catch per tow, zeros included
    mean_CPUE = mean(sample_CPUE), 
    total_CPUE = sum(sample_CPUE),
    .groups = "drop"
  )

```

I'd say where `prop_zero` = 1, these aren't useful, but the next "worst" is `USGSbenthic` (`prop_zero` = 0.590) and that seems super good enough. However, we should probably see how well these sampling efforts have been applied to the areas of interest.

```{r}
#| label: sites
#| echo: false

load(here("data/PrioritySites.RData"))
PrioritySites

# these are, presumably, the data from the relevant sites
df <- Bugs_allfilters %>% filter(Project_na %in% PrioritySites$Project_na)

# turns out that Bugs_allfilters already includes only the priority sites

```

```{r}
#| label: data-selection


# remove those projects w a higher proportion (>=0.75) of 0 insect catches
insect_summary <-
  Bugs_allfilters %>%
  filter(Class == "Insecta") %>%
  group_by(Project_na, Source, TowType, SampleID) %>%
  summarise(sample_CPUE = sum(CPUE, na.rm = TRUE), .groups = "drop") %>% # one row per tow (sample unit)
  group_by(Project_na, Source, TowType) %>%
  summarise(
    n_samples = n_distinct(SampleID),
    n_zero_samples = sum(sample_CPUE == 0), # tows that caught zero insects
    prop_zero = mean(sample_CPUE == 0), # handy for the hurdle question
    mean_CPUE = mean(sample_CPUE), # mean catch per tow, zeros included
    total_CPUE = sum(sample_CPUE),
    .groups = "drop"
  ) %>% 
  filter(prop_zero < 0.75)

insect_summary %>%
  group_by(Source, TowType) %>%
  summarise(
    overall_mean_CPUE = mean(mean_CPUE),
    median_mean_CPUE  = median(mean_CPUE),
    mean_prop_zero    = mean(prop_zero),
    n_sites           = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(overall_mean_CPUE))

# all sites in the filtered dataset
all_sites <- insect_summary %>%
  distinct(Project_na) %>% 
  print()

# sites that were sampled with FRP + SAV
frp_sav_sites <- insect_summary %>%
  filter(Source == "FRP", TowType == "SAV") %>%
  distinct(Project_na) %>% 
  print()

# sites that were NOT sampled with FRP + SAV
all_sites %>%
  anti_join(frp_sav_sites, by = "Project_na")

```

# FRP + SAV dataset

Build the modelling dataset: FRP + SAV tows only, insect CPUE (counts/volume) aggregated to the sample level, retaining covariates for site, inside/outside, and year.

```{r}
#| label: frp-sav-data

# ⚠️ run this first to see all column names and adjust below if needed
# names(Bugs_allfilters)

insect_df <-
  Bugs_allfilters %>%
  filter(
    Source   == "FRP",
    TowType  == "SAV",
    Class    == "Insecta"
  ) %>%
  group_by(
    SampleID,
    Project_na,
    Year,
    Type
  ) %>%
  summarise(
    sample_CPUE = sum(CPUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Year       = factor(Year),
    Type       = factor(Type),
    Project_na = factor(Project_na)
  )

glimpse(insect_df)

# quick sanity check
insect_df %>%
  summarise(
    n_samples    = n(),
    n_sites      = n_distinct(Project_na),
    n_years      = n_distinct(Year),
    prop_zero    = mean(sample_CPUE == 0),
    mean_CPUE    = mean(sample_CPUE),
    median_CPUE  = median(sample_CPUE[sample_CPUE > 0])  # median of non-zeros
  )

```

# distribution of CPUE

Before committing to a hurdle lognormal, check (1) what fraction of samples are zero, (2) whether the non-zero CPUE values are approximately lognormal, and (3) whether zeros or the shape of non-zeros vary meaningfully across sites/years.

```{r}
#| label: cpue-distribution
#| fig-width: 10
#| fig-height: 8

# --- panel A: full distribution including zeros ---
p_all <- ggplot(insect_df, aes(x = sample_CPUE)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  labs(title = "A: All CPUE (including zeros)",
       x = "CPUE (counts/volume)", y = "count") +
  theme_bw()

# --- panel B: non-zero CPUE on log scale ---
p_nonzero <- insect_df %>%
  filter(sample_CPUE > 0) %>%
  ggplot(aes(x = sample_CPUE)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  scale_x_log10() +
  labs(title = "B: Non-zero CPUE (log10 x-axis)",
       x = "CPUE (counts/volume)", y = "count") +
  theme_bw()

# --- panel C: QQ-plot against lognormal ---
# points should fall close to the line if lognormal is appropriate
p_qq <- insect_df %>%
  filter(sample_CPUE > 0) %>%
  mutate(log_CPUE = log(sample_CPUE)) %>%
  ggplot(aes(sample = log_CPUE)) +
  stat_qq(colour = "steelblue", alpha = 0.4) +
  stat_qq_line(colour = "firebrick") +
  labs(title = "C: QQ-plot of log(CPUE) vs normal",
       x = "theoretical quantiles", y = "sample quantiles") +
  theme_bw()

# --- panel D: proportion zeros by site ---
p_zeros <- insect_df %>%
  group_by(Project_na) %>%
  summarise(prop_zero = mean(sample_CPUE == 0), n = n(), .groups = "drop") %>%
  ggplot(aes(x = reorder(Project_na, prop_zero), y = prop_zero)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "D: Proportion zeros by site (n = samples)",
       x = NULL, y = "prop. zero") +
  theme_bw()

(p_all | p_nonzero) / (p_qq | p_zeros)

```

```{r}
#| label: zeros-by-year-type

# proportion zeros broken down by year and inside/outside
insect_df %>%
  group_by(Year, Type) %>%
  summarise(
    n          = n(),
    prop_zero  = mean(sample_CPUE == 0),
    mean_CPUE  = mean(sample_CPUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Year, y = prop_zero, colour = Type, group = Type)) +
  geom_line() +
  geom_point(aes(size = n)) +
  labs(title = "Proportion zero catches by year and type",
       y = "prop. zero", colour = "Type", size = "n samples") +
  theme_bw()

```

# hurdle models

We fit a series of hurdle lognormal models of increasing complexity. All models include a random intercept for `Year` to account for inter-annual variation (as in the bivalve analysis). The hurdle component (probability of zero catch) is modelled separately; we start with a fixed hurdle intercept and can relax that later.

Models fitted:

| model | fixed effects (mu)     | fixed effects (hu) |
|-------|------------------------|--------------------|
| `m0`  | intercept only         | intercept only     |
| `m1`  | Type (in/out)          | intercept only     |
| `m2`  | Project_na             | intercept only     |
| `m3`  | Type + (1\|Project_na) | intercept only     |
| `m4`  | Type + (1\|Project_na) | Type               |

Model `m4` is the most flexible baseline: it includes a site-level random intercept, allows the non-zero mean to vary by inside/outside, and lets the hurdle probability differ by type.

```{r}
#| label: hurdle-priors

# brms treats the intercept as class = Intercept (separate from class = b),
# so we split priors into a base set valid for all models and additive sets
# for models with fixed effects or a modelled hu component.

# valid for every model
priors_base <- c(
  prior(normal(5, 3),   class = Intercept),            # mu intercept (log scale)
  prior(normal(0, 1),   class = Intercept, dpar = hu), # hu intercept (logit scale)
  prior(exponential(1), class = sd),                   # random effect SDs
  prior(exponential(1), class = sigma)                 # lognormal residual SD
)

# add to models that have mu fixed effects (m1–m4)
prior_b    <- prior(normal(0, 1), class = b)

# add to models that have hu fixed effects (m4 only)
prior_b_hu <- prior(normal(0, 1), class = b, dpar = hu)

```

```{r}
#| label: m0-intercept-only
#| cache: true

m0 <- brm(
  bf(sample_CPUE ~ 1 + (1 | Year),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior   = priors_base,
  data    = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores   = 4,
  seed    = 42,
  file    = here("outputs/m0_insects")  # cache to disk
)
pp_check(m0)
summary(m0)

```

```{r}
#| label: m1-type
#| cache: true

m1 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior   = c(priors_base, prior_b),
  data    = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores   = 4,
  seed    = 42,
  file    = here("outputs/m1_insects")
)
summary(m1)

```

```{r}
#| label: m2-site
#| cache: true

# Project_na as fixed effect — works if samples-per-site are adequate;
# if many sites with few samples each, consider (1 | Project_na) instead
m2 <- brm(
  bf(sample_CPUE ~ Project_na + (1 | Year),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior   = c(priors_base, prior_b),
  data    = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores   = 4,
  seed    = 42,
  file    = here("outputs/m2_insects")
)
summary(m2)

```

```{r}
#| label: m3-type-site-re
#| cache: true

m3 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year) + (1 | Project_na),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior   = c(priors_base, prior_b),
  data    = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores   = 4,
  seed    = 42,
  file    = here("outputs/m3_insects")
)
summary(m3)

```

```{r}
#| label: m4-full-baseline
#| cache: true

# most flexible baseline: Type in both components, site random intercept
m4 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year) + (1 | Project_na),
     hu ~ Type),
  family  = hurdle_lognormal(),
  prior   = c(priors_base, prior_b, prior_b_hu),
  data    = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores   = 4,
  seed    = 42,
  file    = here("outputs/m4_insects")
)
summary(m4)

```

```{r}
#| label: model-comparison

# LOO-CV comparison across all models
m0 <- add_criterion(m0, "loo")
m1 <- add_criterion(m1, "loo")
m2 <- add_criterion(m2, "loo")
m3 <- add_criterion(m3, "loo")
m4 <- add_criterion(m4, "loo")

loo_compare(m0, m1, m2, m3, m4)

```
