---
title: "insects"
author: "Pete Nelson"
format: html
editor: visual
---

# intro

Explore Rosie's data to see what monitoring programs and sampling techniques seem adequate for studying the occurrence of insects. Then, to see how effectively these have been applied to the project areas that we're interested in. I'll start by loading some libraries and then looking at some of the data files that Rosie's provided (no

```{r}
#| label: data
#| echo: false

library(here)
library(janitor)
library(tidyverse)
library(brms)
library(patchwork)

load(here("data/AllWetlandBugs.RData")) # AllBugs
load(here("data/AllWetlandBugs_2010onwards.RData")) # Allbugs_Oct2025
load(here("data/Bugs_allfilters.RData"))
load(here("data/Insects.RData"))
glimpse(AllBugs) # 427,198 x 38
glimpse(Allbugs_Mar2026) # 2,266,598 x 40
glimpse(Bugs_allfilters) # 874,790 x 47
glimpse(Insects) # 7,601 x 27

```

I'm unfamiliar with all of the sites, most of the monitoring programs, and many of the sampling techniques (or at least what their codes refer to). The `Insects` data file is limited to chironomids and all other insects, so I'm going to poke around in `Bugs_allfilters` data (though it may well be that that's the best taxonomic resolution we can hope for). This may be the wrong place to start; if so, I'm hoping it'll be relatively easy to correct later.

# data exploration

Add `brms` and `patchwork` to the library block if not already loaded.

I want to see how well different sampling programs ('Source') and sampling techniques ('TowType') work for insects.

```{r}
#| label: sampling-efficacy

Bugs_allfilters %>%
  filter(Class == "Insecta") %>%
  group_by(Source, TowType, SampleID) %>%
  summarise(sample_CPUE = sum(CPUE, na.rm = TRUE), .groups = "drop") %>% # one row per tow (sample unit)
  group_by(Source, TowType) %>%
  summarise(
    n_samples = n_distinct(SampleID),
    # tows that caught zero insects
    n_zero_samples = sum(sample_CPUE == 0), 
    # handy for the expected hurdle question
    prop_zero = mean(sample_CPUE == 0), 
    # mean catch per tow, zeros included
    mean_CPUE = mean(sample_CPUE), 
    total_CPUE = sum(sample_CPUE),
    .groups = "drop"
  )

```

I'd say where `prop_zero` = 1, these aren't useful, but the next "worst" is `USGSbenthic` (`prop_zero` = 0.590) and that seems super good enough. However, we should probably see how well these sampling efforts have been applied to the areas of interest.

```{r}
#| label: sites
#| echo: false

load(here("data/PrioritySites.RData"))
PrioritySites

# these are, presumably, the data from the relevant sites
df <- Bugs_allfilters %>% filter(Project_na %in% PrioritySites$Project_na)

# turns out that Bugs_allfilters already includes only the priority sites

```

```{r}
#| label: data-selection


# remove those projects w a higher proportion (>=0.75) of 0 insect catches
insect_summary <-
  Bugs_allfilters %>%
  filter(Class == "Insecta") %>%
  group_by(Project_na, Source, TowType, SampleID) %>%
  summarise(sample_CPUE = sum(CPUE, na.rm = TRUE), .groups = "drop") %>% # one row per tow (sample unit)
  group_by(Project_na, Source, TowType) %>%
  summarise(
    n_samples = n_distinct(SampleID),
    n_zero_samples = sum(sample_CPUE == 0), # tows that caught zero insects
    prop_zero = mean(sample_CPUE == 0), # handy for the hurdle question
    mean_CPUE = mean(sample_CPUE), # mean catch per tow, zeros included
    total_CPUE = sum(sample_CPUE),
    .groups = "drop"
  ) %>% 
  filter(prop_zero < 0.75)

insect_summary %>%
  group_by(Source, TowType) %>%
  summarise(
    overall_mean_CPUE = mean(mean_CPUE),
    median_mean_CPUE = median(mean_CPUE),
    mean_prop_zero = mean(prop_zero),
    n_sites = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(overall_mean_CPUE))

# all sites in the filtered dataset
all_sites <- insect_summary %>%
  distinct(Project_na) %>% 
  print()

# sites that were sampled with FRP + SAV
frp_sav_sites <- insect_summary %>%
  filter(Source == "FRP", TowType == "SAV") %>%
  distinct(Project_na) %>% 
  print()

# sites that were NOT sampled with FRP + SAV
all_sites %>%
  anti_join(frp_sav_sites, by = "Project_na")

```

# FRP + SAV dataset

Build the modelling dataset: FRP + SAV tows only, insect CPUE (counts/volume) aggregated to the sample level, retaining covariates for site, inside/outside, and year.

```{r}
#| label: frp-sav-data

# run this first to see all column names and adjust below if needed
# names(Bugs_allfilters)

insect_df <-
  Bugs_allfilters %>%
  filter(
    Source == "FRP",
    TowType == "SAV",
    Class == "Insecta"
  ) %>%
  group_by(
    SampleID,
    Project_na,
    Year,
    Type
  ) %>%
  summarise(
    sample_CPUE = sum(CPUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Year = factor(Year),
    Type = factor(Type),
    Project_na = factor(Project_na)
  )

glimpse(insect_df)

# quick check
insect_df %>%
  summarise(
    n_samples = n(),
    n_sites = n_distinct(Project_na),
    n_years = n_distinct(Year),
    prop_zero = mean(sample_CPUE == 0),
    mean_CPUE = mean(sample_CPUE),
    median_CPUE = median(sample_CPUE[sample_CPUE > 0])  # median of non-zeros
  )

```

# distribution of CPUE

Before committing to a hurdle lognormal, check (1) what fraction of samples are zero, (2) whether the non-zero CPUE values are approximately lognormal, and (3) whether zeros or the shape of non-zeros vary meaningfully across sites/years.

```{r}
#| label: cpue-distribution
#| fig-width: 10
#| fig-height: 8

# --- panel A: full distribution including zeros ---
p_all <- ggplot(insect_df, aes(x = sample_CPUE)) +
  geom_histogram(bins = 60, fill = "lightblue", colour = "white", linewidth = 0.2) +
  labs(title = "A: All CPUE (including zeros)",
       x = "CPUE (counts/volume)", y = "count") +
  theme_bw()

# --- panel B: non-zero CPUE on log scale ---
p_nonzero <- insect_df %>%
  filter(sample_CPUE > 0) %>%
  ggplot(aes(x = sample_CPUE)) +
  geom_histogram(bins = 60, fill = "lightblue", colour = "white", linewidth = 0.2) +
  scale_x_log10() +
  labs(title = "B: Non-zero CPUE (log10 x-axis)",
       x = "CPUE (counts/volume)", y = "count") +
  theme_bw()

# --- panel C: QQ-plot against lognormal ---
# points should fall close to the line if lognormal is appropriate
p_qq <- insect_df %>%
  filter(sample_CPUE > 0) %>%
  mutate(log_CPUE = log(sample_CPUE)) %>%
  ggplot(aes(sample = log_CPUE)) +
  stat_qq(colour = "lightblue", alpha = 0.4) +
  stat_qq_line(colour = "firebrick") +
  labs(title = "C: QQ-plot of log(CPUE) vs normal",
       x = "theoretical quantiles", y = "sample quantiles") +
  theme_bw()

# --- panel D: proportion zeros by site ---
p_zeros <- insect_df %>%
  group_by(Project_na) %>%
  summarise(prop_zero = mean(sample_CPUE == 0), n = n(), .groups = "drop") %>%
  ggplot(aes(x = reorder(Project_na, prop_zero), y = prop_zero)) +
  geom_col(fill = "lightblue") +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "D: Proportion zeros by site (n = samples)",
       x = NULL, y = "prop. zero") +
  theme_bw()

(p_all | p_nonzero) / (p_qq | p_zeros)

```

Several useful things here.

**Panel B and C together** are the most important. The non-zero CPUE on log scale (B) looks remarkably flat/uniform rather than bell-shaped — that's not what I'd expect from a lognormal and the QQ-plot (C) confirms this: the middle of the distribution tracks the line reasonably well, but both tails deviate. A true lognormal should curve upward in the upper tail, not spike like that. This suggests the lognormal assumption is passable but not great, and those high-value outliers are influential.

**Panel D** is striking — Browns, Chipps, Winter, and Ryer all have 25–50% zero catches, while the bottom five sites (Web Tract Berms, LICB, Liberty, Flyway Farms, Decker) have essentially NO zeros. That's a strong site-level pattern in the *hurdle* component that my models aren't capturing because `hu ~ 1` pools everything. The top four sites are almost certainly our outside or less-vegetated sites — worth checking.

**Panel A** just confirms the extreme right skew and those two outlier samples around 1250 and 2000 that are visible above the main mass.

**What I'd do next:** First, ask the group: what's different about the groups in panel D, those with substantial frequencies of zero and those with none? Then, check whether the high-zero sites map onto a particular Type (inside/outside) or characteristic, and consider fitting a model with `hu ~ Project_na` or at minimum `hu ~ Type` to let zero probability vary. That panel D pattern looks like real signal that the current models are ignoring.

```{r}
#| label: zeros-by-year-type

# proportion zeros broken down by year and inside/outside
insect_df %>%
  group_by(Year, Type) %>%
  summarise(
    n = n(),
    prop_zero = mean(sample_CPUE == 0),
    mean_CPUE = mean(sample_CPUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Year, y = prop_zero, colour = Type, group = Type)) +
  geom_line() +
  geom_point(aes(size = n)) +
  labs(title = "Proportion zero catches by year and type",
       y = "prop. zero", colour = "Type", size = "n samples") +
  theme_bw()

```

# exploratory plots

Before modelling, visualise mean CPUE (all samples, zeros included) by site and by inside/outside type, broken down by year to show temporal context.

```{r}
#| label: cpue-by-site
#| fig-width: 9
#| fig-height: 5

site_summary <- insect_df %>%
  group_by(Project_na, Type) %>%
  summarise(
    n = n(),
    mean_CPUE  = mean(sample_CPUE),
    se_CPUE = sd(sample_CPUE) / sqrt(n),
    prop_zero = mean(sample_CPUE == 0),
    .groups = "drop"
  )

ggplot(site_summary,
       aes(x = reorder(Project_na, mean_CPUE), y = mean_CPUE, 
           fill = Type, group = Type)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = mean_CPUE - se_CPUE,
                    ymax = mean_CPUE + se_CPUE),
                width = 0.3,
                position = position_dodge(width = 0.9)) +
  coord_flip() +
  scale_fill_manual(values = c("Inside" = "lightblue", "Outside" = "gray")) +
  labs(title = "Mean insect CPUE by site",
       subtitle = "Error bars = ±1 SE",
       x = NULL, y = "mean CPUE (counts/volume)", fill = "Type") +
  theme_bw()

```

```{r}
#| label: cpue-by-type-year
#| fig-width: 8
#| fig-height: 5

type_year_summary <- insect_df %>%
  group_by(Type, Year) %>%
  summarise(
    n = n(),
    mean_CPUE = mean(sample_CPUE),
    se_CPUE = sd(sample_CPUE) / sqrt(n),
    .groups = "drop"
  )

ggplot(type_year_summary,
       aes(x = Year, y = mean_CPUE, colour = Type, group = Type)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n)) +
  geom_errorbar(aes(ymin = mean_CPUE - se_CPUE,
                    ymax = mean_CPUE + se_CPUE),
                width = 0.2) +
  scale_colour_manual(values = c("Inside" = "green", "Outside" = "blue")) +
  labs(title = "Mean insect CPUE by type and year",
       subtitle = "Point size = n samples; error bars = ±1 SE",
       y = "mean CPUE (counts/volume)", colour = "Type", size = "n") +
  theme_bw()

```

```{r}
#| label: cpue-by-site-year
#| fig-width: 10
#| fig-height: 6

# jittered raw points + site mean per year — shows where the data are sparse
insect_df %>%
  ggplot(aes(x = Year, y = sample_CPUE, colour = Type)) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  stat_summary(fun = mean, geom = "line",
               aes(group = Project_na), linewidth = 0.6) +
  stat_summary(fun = mean, geom = "point", size = 2.5, shape = 18) +
  scale_y_log10(labels = scales::comma) +
  scale_colour_manual(values = c("Inside" = "green", "Outside" = "blue")) +
  facet_wrap(~ Project_na) +
  labs(title = "CPUE by site and year (log scale)",
       subtitle = "Points = individual tows; diamonds = site mean; lines connect means",
       y = "CPUE (log scale)", colour = "Type") +
  theme_bw(base_size = 11)

```

# hurdle models

We fit a series of hurdle lognormal models of increasing complexity. All models include a random intercept for `Year` to account for inter-annual variation (as in the bivalve analysis). The hurdle component (probability of zero catch) is modelled separately; we start with a fixed hurdle intercept and can relax that later.

Models fitted:

| model | fixed effects (mu)     | fixed effects (hu) |
|-------|------------------------|--------------------|
| `m0`  | intercept only         | intercept only     |
| `m1`  | Type (in/out)          | intercept only     |
| `m2`  | Project_na             | intercept only     |
| `m3`  | Type + (1\|Project_na) | intercept only     |
| `m4`  | Type + (1\|Project_na) | Type               |

Model `m4` is the most flexible baseline: it includes a site-level random intercept, allows the non-zero mean to vary by inside/outside, and lets the hurdle probability differ by type.

```{r}
#| label: hurdle-priors

# brms treats the intercept as class = Intercept (separate from class = b),
# so we split priors into a base set valid for all models and additive sets
# for models with fixed effects or a modelled hu component.

# valid for every model
priors_base <- c(
  prior(normal(5, 3), class = Intercept), # mu intercept (log scale)
  prior(normal(0, 1), class = Intercept, dpar = hu), # hu intercept (logit scale)
  prior(exponential(1), class = sd), # random effect SDs
  prior(exponential(1), class = sigma) # lognormal residual SD
)

# add to models that have mu fixed effects (m1–m4)
prior_b <- prior(normal(0, 1), class = b)

# add to models that have hu fixed effects (m4 only)
prior_b_hu <- prior(normal(0, 1), class = b, dpar = hu)

```

```{r}
#| label: m0-intercept-only
#| cache: true

m0 <- brm(
  bf(sample_CPUE ~ 1 + (1 | Year),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior = priors_base,
  data = insect_df,
  chains = 4, iter = 3000, warmup = 1000, thin = 10,
  cores = 4,
  seed = 42,
  file = here("outputs/m0_insects") 
)
pp_check(m0)
plot(m0) # values extend WAAAY out there; model thinks extreme catches plausible?
summary(m0)

```

The model converged cleanly (all Rhat = 1.00, ESS well above 400 for all parameters). Here's what each piece tells us:

**Hurdle component (hu_Intercept = -1.73)**
`plogis(-1.73) ≈ 0.15` — about 15% of FRP+SAV insect samples catch nothing. That's a moderate zero rate, consistent with a hurdle model being appropriate but not dominated by zeros.

**Lognormal component (Intercept = 3.51)**
The expected log-CPUE for a non-zero catch is 3.51, so the median CPUE on the raw scale is `exp(3.51) ≈ 33` insects/volume. The wide CI (2.44–4.69)...year-to-year variability absorbed by the random effect.

**Year random effect (sd = 1.03)**
This is large — on the log scale, a SD of 1 means years may differ by a factor of `exp(1) ≈ 2.7` in median CPUE. With only 4 levels the estimate is uncertain (0.35–2.41), but it's clearly non-trivial. This justifies keeping `(1 | Year)` in all subsequent models.

**sigma = 1.66**
Within-year, within-site residual variation on the log scale is also large. `exp(1.66) ≈ 5.3`, meaning even after accounting for year, individual tow catches vary by roughly 5-fold. Insects are patchy — that's expected.

**Bottom line for m0:** \~15% zeros, median catch \~33/volume, but enormous variability both across years and within years. The year random effect is explaining a lot, and the wide sigma suggests site or type effects (m1–m4) might explain some of that residual spread.

```{r}
#| label: m1-type
#| cache: true

m1 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior = c(priors_base, prior_b),
  data = insect_df,
  chains = 4, iter = 3000, warmup = 1000, thin = 10,
  cores = 4,
  seed = 42,
  file = here("outputs/m1_insects")
)
pp_check(m1) # ok but again expecting lrg numbers we didn't see
plot(m1) # the interesting (in v out) not so interesting
summary(m1)

```

Again, zeros are rare.

```{r}
#| label: m2-site
#| cache: true

# Project_na as fixed effect — works if samples-per-site are adequate;
# if many sites with few samples each, consider (1 | Project_na) instead
m2 <- brm(
  bf(sample_CPUE ~ Project_na + (1 | Year),
     hu ~ 1),
  family = hurdle_lognormal(),
  prior = c(priors_base, prior_b),
  data = insect_df,
  control = list(adapt_delta = 0.95),
  # had 3 diverg transitions, 
  # so increased adapt_delta, iter and warmup and re-ran
  chains  = 4, iter = 4000, warmup = 1500, thin = 10,
  cores = 4,
  seed = 42,
  file_refit = "on_change", # "always" or "on_change"
  file    = here("outputs/m2_insects")
)
pp_check(m2)
plot(m2)
summary(m2)

```

```{r}
#| label: m3-type-site-re
#| cache: true

m3 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year) + (1 | Project_na),
     hu ~ 1),
  family  = hurdle_lognormal(),
  prior = c(priors_base, prior_b),
  data = insect_df,
  chains  = 4, iter = 3000, warmup = 1000, thin = 10,
  cores = 4,
  seed = 42,
  file = here("outputs/m3_insects")
)
pp_check(m3)
plot(m3)
summary(m3)

```

```{r}
#| label: m4-full-baseline
#| cache: true

# most flexible baseline: Type in both components, site random intercept
m4 <- brm(
  bf(sample_CPUE ~ Type + (1 | Year) + (1 | Project_na),
     hu ~ Type),
  family  = hurdle_lognormal(),
  prior = c(priors_base, prior_b, prior_b_hu),
  data = insect_df,
  control = list(adapt_delta = 0.95),
  # had 3 diverg transitions, 
  # so increased adapt_delta, iter and warmup and re-ran
  chains  = 4, iter = 4000, warmup = 1500, thin = 10,
  cores = 4,
  seed = 42,
  file_refit = "on_change", # "always" or "on_change"
  file = here("outputs/m4_insects")
)
pp_check(m4)
plot(m4)
summary(m4)

```

```{r}
#| label: model-comparison

# LOO-CV comparison across all models
m0 <- add_criterion(m0, "loo")
m1 <- add_criterion(m1, "loo")
# observations w pareto_k > 0.7...
m2 <- add_criterion(m2, "loo", moment_match = TRUE)
m3 <- add_criterion(m3, "loo", moment_match = TRUE)
m4 <- add_criterion(m4, "loo", moment_match = TRUE)

loo_compare(m0, m1, m2, m3, m4)

```

# status

**Neither inside/outside nor site identity is explaining much of the variance in insect CPUE**, at least not beyond what year already captures.

The large sigma (1.66) and year SD (∼1.0) from m0 are still doing most of the work. With only 75 samples across 4 years, we seem to be underpowered to detect site or type effects even if they exist: **the data don't show a clear signal for inside vs. outside, or for site differences, beyond what year-to-year variation already explains.**
