---
title: "join data"
author: "Bobbie Flores, Data Stewarship Manager, Cramer Fish Sciences"
date: "2025-05-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction:

This markdown is to help data users join the DWR data collected by Cramer Fish Sciences at Dutch Slough from 2021-2023 using the Sampling Platform methodoloy.

Files included in the final data package include:
        - Metadata that links all datasets together
                "DS_metadata_2021-2032.csv"
        - Water quality data recorded from sonde
                "DS_waterquality_2021-2023.csv"
        - Zooplankton observations from recorded images
                "DS_zoops_2021-2023.csv"
        - Fish observations from recorded video
                "DS_fish_2021-2023.csv"
        - Table of included data in DWR format
                "Food web datasets.xlsx"
        - Readme with explanations of all data fields
                "Readme.xlsx"
        - This R Markdown file
                "join data.Rmd"
                
*All tables join on the field "skeleton"*   

NA data should be assumed to have not been collected on that day/transect, corrupt, or fell outside of acceptable QC bounds and was removed.

                
If there are any questions about the dataset or this markdown, please contact the author at bobbie.flores@fishsciences.net

# Libraries:
There is no need to load special packages for this code as originally written.

# Load data
```{r}
metadata<-read.csv("DS_metadata_2021-2023.csv")
waterquality<-read.csv("DS_waterquality_2021-2023.csv")
zoop<-read.csv("DS_zoops_2021-2023.csv")
fish<-read.csv("DS_fish_2021-2023.csv")
```

# Join data

## Water quality
```{r}
wq<-merge(waterquality, metadata)
```

## Zoops
```{r}
z1<-merge(zoop, metadata)
```

## Fish
```{r}
f1<-merge(fish, metadata)
```



